---
title: "From rMATS output to heatmap of diff splicing events"
author: "Maria Kondili"
date: "12/04/2021"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(glue))

```

Objective: Use rMATS output & Merge all events-types & Create heatmap of diff splicing events

```{r}

suppressPackageStartupMessages(library(dplyr))

work_dir="/projects/mbnl_dct/mouse_seq/rMATS_splicingCounts/"
#dir.create(work_dir, showWarnings = F)

Decoy_vs_Saline_dir <- paste0(work_dir,"MBNLd_vs_Saline_Splicing/")
# Decoy_vs_Saline_dir <- paste0(work_dir,"Encode_MBNLd_vs_Saline_Splicing/")

AavGfp_vs_Saline_dir <- paste0(work_dir,"AAVGFP_vs_Saline_Splicing/")

```

Alignment done with default params, as for Human-cells (with STAR)

Create output folder to save merged table

```{r}
outdir.dct <- paste0(work_dir,"modif_MBNLd_vs_Saline_Splicing/")
outdir.gfp <- paste0(work_dir,"modif_AAVGFP_vs_Saline_Splicing/")
##! ->hadn't run AAVGFP-vs-Saline rMATS with STAR !>done June 2021
dir.create(outdir.dct, showWarnings = F)
dir.create(outdir.gfp, showWarnings = F)
```


```{r}

source("add_exonLength_and_Coord.R")

splice_events <- c("A3SS", "A5SS","MXE", "RI", "SE" )
## output in same folder but with suffix "_exonCoord"


for (s in splice_events) {
  cat(" Launching event :", s,"\n")
  # An rMATS output is read : read.delim(paste0(Decoy_vs_Saline_dir, s, ".MATS.JCEC.txt") ,as.is=TRUE,header=TRUE)
  # add_exonLen_and_coord(splice_event=s, tab_suffix=".MATS.JCEC.txt",
  #                       in_dir=Decoy_vs_Saline_dir,
  #                       out_dir= outdir.saline)
  #
  add_exonLen_and_coord(splice_event=s, tab_suffix=".MATS.JCEC.txt",
                        in_dir=Decoy_vs_Saline_dir,
                        out_dir= outdir.dct)

  add_exonLen_and_coord(splice_event=s, tab_suffix=".MATS.JCEC.txt",
                        in_dir=AavGfp_vs_Saline_dir,
                        out_dir= outdir.gfp)

}

glue("New tables with modified columns are created:\n {list.files(outdir.gfp)}")
glue("New tables with modified columns are created:\n {list.files(outdir.dct)}")

```

> call function for : MBNLdecoy-vs-Saline

```{r}
new_suffix = ".JCEC_PerRepl_exonCoord.txt"

colsOI <- c("GeneID",	"geneSymbol","chr","strand",
            "IncFormLen", "SkipFormLen","ExonLength",
            "IncLevel1_r1","IncLevel1_r2","IncLevel1_r3",
            "IncLevel2_r1","IncLevel2_r2","IncLevel2_r3",
            "splice_event", "IncLevelDifference",
            "alternative_exon_coordinates","PValue","FDR" )

source("merge_spliceTypes_and_save.R")

ase.dct.star <- merge_spliceTypes_and_save(splice_events,
                                             new_suffix, colsOI,
                                             outdir=outdir.dct,
                                             filename="Saline_vs_MBNLdecoy")

ase.gfp.star <- merge_spliceTypes_and_save(splice_events,
                                             new_suffix, colsOI,
                                             outdir=outdir.gfp,
                                             filename="Saline_vs_AAVGFP")


```


```{r,echo=F}

cat("Merged Splicing Events for Saline-vs-MBNLd are in file :'merged_modif_ASE_Saline_vs_MBNLdecoy.tsv' ")

cat("Merged Splicing Events for Saline-vs-AAVGFP are in file :'merged_modif_ASE_Saline_vs_AAVGFP.tsv' ")

```


```{r}

ase.dct.star <- mutate(ase.dct.star ,
                     "dPSI" = IncLevelDifference )

ase.gfp.star <- mutate(ase.gfp.star ,
                     "dPSI" = IncLevelDifference )


```

```{r plot_heatmap_function}

plot_heatmap_signif_splice <- function(ase, fdr.co, psi.co, pair, inc_level_cols= c(8:13), plot_name) {

    ase.filt <- subset(ase, (FDR < fdr.co &  abs(dPSI) > psi.co) )


    ###  z.gene =  (E(g) - Mean(g.replicates) ) / M.A.D (g.replicates), mad=median.absolute.deviation
    x <- ase.filt[,inc_level_cols]
    s1 = strsplit(pair, "_vs_")[[1]][1]
    s2 = strsplit(pair, "_vs_")[[1]][2]
    colnames(x) <- c( paste(s1,"_", c("r1","r2","r3"),sep=""),
                      paste(s2,"_",c("r1","r2","r3"),sep=""))


    cal_z_score <- function(x){
      (x - mean(x)) / sd(x)
    }
    x_norm <- t(apply(x, 1, cal_z_score))
    # x_norm <- x_norm[-which(apply(x_norm, 2, is.na)) ,]

    #Rowv=as.dendrogram(hcl_row),tree=ct))
    library(gplots)
    library(RColorBrewer)


    heatmap.2(x_norm,
              Rowv=TRUE,
              Colv=F,
              dendrogram ='row',
              scale="row", na.rm=TRUE,
              trace='none',
              labCol=colnames(x),
              col=rev(colorRampPalette(brewer.pal(10, "RdBu"))(256)),
              margins=c(9,2),
              cexCol=1.3,
              srtCol=45,
              labRow="",
              offsetCol=0,
              key=TRUE,
              keysize = 1.5,
              density.info="density",
              key.title="Color range",
              key.xlab="norm.counts",
              key.ylab="Counts",
              key.par=list(mar=c(3,4,3,1)),
              notecex=3,
              main=paste(nrow(x_norm),"signif. Diff/ial splicing events\nin ",pair))
    #print(hmp)
    return(x)

}

```

Call for ASE:Saline-vs-MBNLdecoy

```{r call function plot}
## FILTER NON-SIGNIFICANT

pair="Saline_vs_MBNLdecoy"
plot_name="Plots/Heatmap_SignifAltSplicing_Encode_Saline_vs_MBNLdecoy.png"

# png(plot_name)
ase.dct.filt.20 <- plot_heatmap_signif_splice(ase.dct.star, fdr.co= 0.05, psi.co=0.20, pair,inc_level_cols= c(8:13), plot_name)
#dev.off()

cat("Genes spliced: \n ")
ase.dct.star[rownames(ase.dct.filt.20),"geneSymbol" ]


```

14 events when psi.co = 0.20

```{r}

ase.dct.filt.15 <- plot_heatmap_signif_splice(ase.dct.star, fdr.co= 0.05, psi.co=0.15, pair, inc_level_cols= c(8:13),plot_name)

cat("Genes spliced: \n ")
ase.dct.star[rownames(ase.dct.filt.15),"geneSymbol" ]


```

18 events when psi.co = 0.15

####--> with AAVGFP-vs-Saline :
```{r}

pair="Saline_vs_AAVGFP"
plot_name="Plots/Heatmap_SignifAltSplicing_Saline_vs_AAVGFP.png"

# png(plot_name)
ase.gfp.filt.20 <- plot_heatmap_signif_splice(ase.gfp.star, fdr.co= 0.05, psi.co=0.20, pair,inc_level_cols=c(8:13), plot_name)

cat("Genes spliced :\n")
cat(ase.gfp.star[rownames(ase.gfp.filt.20),'geneSymbol'])

ase.gfp.filt.15 <- plot_heatmap_signif_splice(ase.gfp.star, fdr.co= 0.05, psi.co=0.15, pair,inc_level_cols=c(8:13), plot_name)

cat(ase.gfp.star[rownames(ase.gfp.filt.15),'geneSymbol'])

```

Find Common Significantly Spliced Genes & events in Mbnldecoy/Saline & AAVGFP/Saline :

```{r}
cat("Genes:\n")

intersect (ase.gfp.star[rownames(ase.gfp.filt.20),"geneSymbol"] ,
           ase.dct.star[rownames(ase.dct.filt.20),"geneSymbol"] )

cat("Events:\n")
intersect(ase.dct.star[rownames(ase.dct.filt.20),"alternative_exon_coordinates"],
          ase.gfp.star[rownames(ase.gfp.filt.20),"alternative_exon_coordinates"] )


```

#### PIE-CHART of SIGNIF-SPLICED EVENTS and Others
Show that differential splicing events in MBNLdecoy compaired to Saline are very few

```{r create_piechart_function}

create_pieChart_Splicing <- function(ase, pair, signif.co,psi.co) {
  ##""" Test with : ase = ase.saline; pair="Saline_vs_MBNLdecoy" ; signif.co =0.05 ; psi.co=0.20 """

  library(ggplot2)

  ## Count events Signif. splicing :
  ## count InclDifference(dPSI) = mean(Saline) -  mean(MBNLdecoy)
  ## INCLUDED : PSI < -0.15
  ## EXCLUDED : PSI > 0.15

  signif <- nrow(subset(ase, (FDR < signif.co  &  abs(IncLevelDifference) >= psi.co )))
  all_splice <-  nrow(ase)

  splice_counts <- data.frame(category=c("All", "Signif & highPSI"),
                              count=c(all_splice,signif))

  ## calculate percentage of each count :
  splice_counts$fraction = round(splice_counts$count / sum(splice_counts$count),2) * 100

  # Compute the cumulative percentages (top of each rectangle)
  splice_counts$ymax = cumsum(splice_counts$fraction)

  # Compute the bottom of each rectangle
  splice_counts$ymin = c(0, head(splice_counts$ymax, n=-1))

  # Compute label position
  # pos <- (splice_counts$ymax + splice_counts$ymin) / 2 # = 49.0 98.5 99.5
  # 98 & 99 are very close, I ll replace them with 97 and 10, left and right of the end of circle.
  splice_counts$labelPosition <- c(50.0, 97.0)

  # Compute a good label
  splice_counts$label <- paste0(splice_counts$category,"\n", splice_counts$fraction, " %")

  # Make the plot
  dir.create("Plots/",showWarnings = F)
  #png(paste0("Plots/Donut_Plot_Signif_SplicingEvents_",pair,".png"))
    ggplot(splice_counts, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=category)) +
      geom_rect() +
      geom_label( x=3.5, aes(y=labelPosition, label=label), size=3) +
      coord_polar(theta="y") +
      xlim(c(2, 4)) +
      theme_void()
 #dev.off()

}

```

call it with ASE table :

```{r, call_function}

create_pieChart_Splicing(ase.dct.star, pair="Saline_vs_MBNLdecoy",signif.co=0.05, psi.co=0.20)

```

####-- with AAVGFP ...

```{r}
create_pieChart_Splicing(ase.gfp.star, pair="Saline_vs_MBNLdecoy",signif.co=0.05, psi.co=0.20)
```
